An efficient approach to approximating the pair distribution function of the inhomogeneous hard-sphere fluid

### 2014-09-02 papers/pair-correlation/paper.tex
finish another response to referee Tested on bennet. Took 0:19.68.

\newcommand{\fixme}[1]{\red{[#1]}}
\newcommand{\newstuff}[1]{\blue{#1}}

@@

Early approximations to the pair distribution function used the
density at one or two positions to determine the pair distribution
function.  There are three common approaches:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) &\approx
  g\left(r_{12}; n\left(\frac{\rr_1+\rr_2}{2}\right)\right)
     & \text{midpoint} \\
  g^{(2)}(\rr_1,\rr_2) &\approx
  g\left(r_{12}; \frac{n(\rr_1)+n(\rr_2)}{2}\right)
     & \text{mean density} \\
  g^{(2)}(\rr_1,\rr_2) &\approx
  \frac{g(r_{12};n(\rr_1))+g(r_{12};n(\rr_2))}{2}
     & \text{mean function}
\end{align}
These approaches have been successfully and widely used in treating
the surface tension of simple fluids~\cite{pressing2003surface,
  salter2008statistical, bongiorno1975modified,
  toxvaerd1976hydrostatic, kalos1977structure, carey2008gradient,
  osborn1980monotonic, mccoy1981comparison, barrett2006some}.  The
mean density approximation has also been quoted (as a goal) by recent
papers that proceed to make further
approximations~\cite{gloor2007prediction, gross2009density}.  However,
these approximations fail dramatically when applied to strongly
inhomogeneous systems such as a dense fluid at a solid surface.  Such
systems exhibit a strongly oscillatory density distribution, with
density peaks that can have local packing fractions greater than
unity, which cannot occur in the bulk reference system that defines
$g(r; n)$.  The above papers restrict themselves to the liquid-vapor
interface, which does not exhibit this pathology.

@@

The CVA requires the radial distribution function expressed as a
function of $r$ and $g_\sigma$.
%
%% \fixme{One approach to constructing $g(r;g_\sigma)$ is to compute $n$
%%   from $g_\sigma$ (using e.g. the Carnahan-Starling formula for
%%   $g_\sigma$) and then find $g(r;n)$ using either some approximate
%%   functional form (e.g. Percus-Yevick) or Monte Carlo simulation
%%   results.}
%
We construct a functional form for $g(r, g_\sigma)$ that allows for
improved computational efficiency.  We introduce the general form that
allows for this efficiency in Section~\ref{sec:efficient}, and we
detail our specific approximation for $g(r,g_\sigma)$ that uses this
general form in Section~\ref{sec:separable-fit}.
%
%% When evaluating the approximations of earlier works described in
%% Section~\ref{sec:gn} \newstuff{for comparison}, we use interpolated
%% Monte Carlo results for $g(r;n)$.

@@

\section{Making the CVA efficient}
\label{sec:efficient}
The existing approaches to approximating the pair distribution
function outlined in Section~\ref{sec:gn} have not been widely used in
the construction of density functionals based on thermodynamic
perturbation theory, largely due to their computational complexity.
While our CVA provides only an incremental improvement in accuracy,
its construction enables significant gains in computational
efficiency, allowing for practical application in density functionals.
We achieve this gain by developing a \emph{separable} fit to the radial
distribution function of the hard-sphere fluid (see
Section~\ref{sec:separable-fit} for details).  This separable fit is
of the form
\begin{align}
  g(r; g_\sigma) &= \sum_{i} a_i(r) b_i(g_\sigma)
\end{align}
where the notable aspect is that the radial distribution function is
written as a sum of terms that are each a simple product of a function
of radius with a function of $g_\sigma$.  This enables us to write
integrals---such as Eq.~\ref{eq:mean-field}---that are linear in the
pair distribution function as a summation of fixed-kernel
convolutions, which may be efficiently computed using Fast Fourier
Transforms (FFTs) in $\mathcal{O}(N\log N)$ time.

@@

We begin by examining the lowest-order correction in TPT, given by
Eq.~\ref{eq:mean-field}.  The two terms that are averaged in
Eq.~\ref{eq:g2-our-mean} give equal contributions to the integral
\begin{align}
  F_1^{\textit{CVA}} &= \tfrac12 \!\! \iint \!\!
  g(r_{12};g_\sigma(\rr_2))n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|)
  d\rr_1d\rr_2.
\end{align}
When we introduce the separable form for $g(r_{12};g_\sigma)$ we can
further simplify this integral as
\begin{align}
  F_1^{\textit{CVA}} \! = \!
  \sum_i \!\tfrac12 \!\! \int \!\! n(\rr_1) \!\!
                            \int \!\!\!
                            %\big(
                            a_i(r_{12})\Phi(r_{12})
                            %\big)
                            %\big(
                            b_i(
g_\sigma(\rr_2))n(\rr_2)
                            %\big)
  d\rr_2d\rr_1
\end{align}
where the functional is written as a summation of integrals of simple
convolutions in three dimensions.  Thus,
each of these integrals may be computed in $\mathcal{O}(N\log N)$
time, which is the same scaling behavior that is common to the most
common and accurate classical density functional theories, most
notably Fundamental Measure Theory (FMT)~\cite{rosenfeld1989free,
  roth2002whitebear}.

>>>>

\newcommand{\fixme}[1]{\red{[#1]}}
\newcommand{\newstuff}[1]{{\color{blue} #1}}

@@

Early approximations to the pair distribution function used the
density at one or two positions to determine the pair distribution
function.  There are three common approaches:
\begin{align}
  g^{(2)}(\rr_1,\rr_2) &\approx
  g\left(r_{12}; n\left(\frac{\rr_1+\rr_2}{2}\right)\right)
     & \text{midpoint} \\
  g^{(2)}(\rr_1,\rr_2) &\approx
  g\left(r_{12}; \frac{n(\rr_1)+n(\rr_2)}{2}\right)
     & \text{mean density} \\
  g^{(2)}(\rr_1,\rr_2) &\approx
  \frac{g(r_{12};n(\rr_1))+g(r_{12};n(\rr_2))}{2}
     & \text{mean function}\label{eq:mean-function}
\end{align}
These approaches have been successfully and widely used in treating
the surface tension of simple fluids~\cite{pressing2003surface,
  salter2008statistical, bongiorno1975modified,
  toxvaerd1976hydrostatic, kalos1977structure, carey2008gradient,
  osborn1980monotonic, mccoy1981comparison, barrett2006some}.  The
mean density approximation has also been quoted (as a goal) by recent
papers that proceed to make further
approximations~\cite{gloor2007prediction, gross2009density}.  However,
these approximations fail dramatically when applied to strongly
inhomogeneous systems such as a dense fluid at a solid surface.  Such
systems exhibit a strongly oscillatory density distribution, with
density peaks that can have local packing fractions greater than
unity, which cannot occur in the bulk reference system that defines
$g(r; n)$.  The above papers restrict themselves to the liquid-vapor
interface, which does not exhibit this pathology.

@@

The CVA requires the radial distribution function expressed as a
function of $r$ and $g_\sigma$.
%
%% \fixme{One approach to constructing $g(r;g_\sigma)$ is to compute $n$
%%   from $g_\sigma$ (using e.g. the Carnahan-Starling formula for
%%   $g_\sigma$) and then find $g(r;n)$ using either some approximate
%%   functional form (e.g. Percus-Yevick) or Monte Carlo simulation
%%   results.}
%
We construct a functional form for $g(r, g_\sigma)$ that allows for
improved computational efficiency.  We introduce the general form that
allows for this efficiency in Section~\ref{sec:efficient}, and we
detail our specific approximation for $g(r,g_\sigma)$ that uses this
general form in Section~\ref{sec:separable-fit}.
%
%% When evaluating the approximations of earlier works described in
%% Section~\ref{sec:gn} for comparison, we use interpolated
%% Monte Carlo results for $g(r;n)$.

@@

\section{Making the CVA efficient}
\label{sec:efficient}
The existing approaches to approximating the pair distribution
function outlined in Section~\ref{sec:gn} have not been widely used in
the construction of density functionals based on thermodynamic
perturbation theory, largely due to their computational complexity.
While our CVA provides only an incremental improvement in accuracy,
its construction enables significant gains in computational
efficiency, allowing for practical application in density functionals.
We achieve this gain by developing a \emph{separable} fit to the radial
distribution function of the hard-sphere fluid (see
Section~\ref{sec:separable-fit} for details).  This separable fit is
of the form
\begin{align}
  g(r; g_\sigma) &= \sum_{i} a_i(r) b_i(g_\sigma)
\end{align}
where the notable aspect is that the radial distribution function is
written as a sum of terms that are each a simple product of a function
of radius with a function of $g_\sigma$.  This enables us to write
integrals---such as Eq.~\ref{eq:mean-field}---that are linear in the
pair distribution function as a summation of fixed-kernel
convolutions, which may be efficiently computed using Fast Fourier
Transforms (FFTs).

\newcommand\Vcell{V_{\textit{cell}}}
\newcommand\Vinteraction{V_{\Phi}}

Computation of the free energy correction from Eq.~\ref{eq:mean-field}
for a periodic system by direct integration requires a nested
integration over the volume of the system~$\Vcell$, and the volume
over which the interaction is nonzero~$\Vinteraction$.  Thus the cost
of computation scales as
$\mathcal{O}\left(\frac{\Vcell\Vinteraction}{\Delta V^2}\right)$ where
$\Delta V$ is the volume resolution of the computational grid.  Direct
integration is the most efficient algorithm when using the existing
functionals for $g^{(2)}(\rr_1,\rr_2)$ described in
Section~\ref{sec:gn}.  The one exception is the ``mean-function''
approximation (Eq.~\ref{eq:mean-function}), which could in principle
be made more efficient using the same technique we describe here.
Because the CVA allows the integral in Eq.~\ref{eq:mean-field} to be
written as a sum of fixed-kernel convolutions, it can be computed
without a nested integral, at the cost of performing a few FFTs.  This
approach scales as $\mathcal{O}\left(\frac{\Vcell}{\Delta
  V}\log\frac{\Vcell}{\Delta V}\right)$, as do most widely used DFT
functionals such as FMT~\cite{rosenfeld1989free, roth2002whitebear}.
%
With this scaling, when examining systems with long interaction
distances or high resolution---which is often necessary when working
with hard-sphere functionals---the CVA has the potential to be far
more efficient than existing methods.

@@

To see how we obtain this improved scaling, we examine the
lowest-order correction in TPT, given by Eq.~\ref{eq:mean-field}.  The
two terms that are averaged in Eq.~\ref{eq:g2-our-mean} give equal
contributions to the integral
\begin{align}
  F_1^{\textit{CVA}} &= \tfrac12 \!\! \iint \!\!
  g(r_{12};g_\sigma(\rr_2))n(\rr_1)n(\rr_2)\Phi(|\rr_1-\rr_2|)
  d\rr_1d\rr_2.
\end{align}
When we introduce the separable form for $g(r_{12};g_\sigma)$ we can
further simplify this integral as
\begin{align}
  F_1^{\textit{CVA}} \! = \!
  \sum_i \!\tfrac12 \!\! \int \!\! n(\rr_1) \!\!
                            \int \!\!\!
                            %\big(
                            a_i(r_{12})\Phi(r_{12})
                            %\big)
                            %\big(
                            b_i(
g_\sigma(\rr_2))n(\rr_2)
                            %\big)
  d\rr_2d\rr_1
\end{align}
where the functional is written as a summation of integrals of simple
convolutions in three dimensions.  Thus, each of these integrals may
be computed in $\mathcal{O}(N\log N)$ time, where $N$ is the number of
grid points in the computational cell.  This is the same scaling as is
required to compute the fundamental measures such as $n_3$ which are
used in FMT.
